---
title: "README.Rmd"
author: "bea-ds"
date: "10/4/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
This document explains how all of the scripts work and how they are connected.

Particularly, this project analyse Human Activity Recognition Using Smartphones Dataset from [1]

[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. 
Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support 
Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). 
Vitoria-Gasteiz, Spain. Dec 2012

There is an run_analysis.R script and does the following:

1. Merging the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement.
3. Uses descriptive activity names to name the activities in the data set
4. Appropriately labels the data set with descriptive variable names.
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

First, we load needed libraries:

```{r}
library(plyr)
library(dplyr)
library(reshape2)
```

Then, we create a data directory if it does not exist:

```{r}
# Creating data directory
if (!file.exists("./data")){ dir.create("./data")}
```
After that, we download the data set:

```{r}
# Downloading data set
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata/projectfiles/UCI%20HAR%20Dataset.zip"
temporalFile <-  tempfile()
download.file(fileUrl, temporalFile)
unzip(temporalFile, exdir = "./data")
unlink(temporalFile)
rm(fileUrl, temporalFile)

```

Now, we read test, train and labels files:

```{r}
# Reading test files
subjectTest <- read.delim ("./data/UCI HAR Dataset/test/subject_test.txt", header=FALSE, col.names = c("subjectId"))
testActivityLabels <- read.delim ("./data/UCI HAR Dataset/test/y_test.txt", header=FALSE, col.names = c("activityLabel"))
testSet <- read.delim ("./data/UCI HAR Dataset/test/X_test.txt", header=FALSE, sep = "")

# Reading train files
subjectTrain <- read.delim ("./data/UCI HAR Dataset/train/subject_train.txt", header=FALSE, col.names = c("subjectId"))
trainActivityLabels <- read.delim ("./data/UCI HAR Dataset/train/y_train.txt", header=FALSE, col.names = c("activityLabel"))
trainSet <- read.delim ("./data/UCI HAR Dataset/train/X_train.txt", header=FALSE, sep = "")

# Reading labels files
activityLabels <- read.delim ("./data/UCI HAR Dataset/activity_labels.txt", header=FALSE, sep = "", col.names = c("activityLabel", "activityName"))
features <- read.delim ("./data/UCI HAR Dataset/features.txt", header=FALSE, sep = "", col.names = c("variableLabel", "variableName"))
```

1: Merging the training and the test sets to create one data set.

First, we proceed to join rows for each train and test file type:

```{r}
# Joining rows for each train and test file type
totalSubjects <- rbind (subjectTrain, subjectTest)
totalActivityLabels <- rbind (trainActivityLabels, testActivityLabels)
totalSet <- rbind (trainSet, testSet)

# Removing data not longer needed
rm (subjectTrain, trainActivityLabels,trainSet, subjectTest, testActivityLabels, testSet)
```

2. Extracts only the measurements on the mean and standard deviation for each measurement.

First, we extract mean() and std() index and values from features data set, and then from the data set:

```{r}
# Extracting mean() and std() index and values from features data set
measures <- grep ("mean\\(\\)|std\\(\\)", features[,2])
measuresNames <- grep ("mean\\(\\)|std\\(\\)", features[,2], value=TRUE)

# Removing feature data set, not longer needed
rm(features)

# Extracting mean() and std() measures from total data set
totalSet <- totalSet[,measures]
names(totalSet) <- measuresNames
```

Then, we join subjects, activities and mean and standard deviation to create one data set:

```{r}

# Joining subjects, activities and mean and standard deviation to create one data set
totalSet <- cbind(totalSubjects, totalActivityLabels,totalSet)

# Removing data not longer needed
rm (totalSubjects, totalActivityLabels, measures, measuresNames)
```

3. Uses descriptive activity names to name the activities in the data set

```{r}
# Converting to character subjectId and activityLabel
totalSet$subjectId <- as.character(totalSet$subjectId)
totalSet$activityLabel <- as.character(totalSet$activityLabel)

# Converting to character activityLabel and activityName
activityLabels$activityLabel <- as.character(activityLabels$activityLabel)
activityLabels$activityName <- as.character(activityLabels$activityName)

# Using descriptive activity names to name the activities in the data set
totalSet$activityLabel <- mapvalues(totalSet$activityLabel, activityLabels$activityLabel, activityLabels$activityName, warn_missing = TRUE)

# Removing data not longer needed
rm (activityLabels)
```
4. Appropriately labels the data set with descriptive variable names:

```{r}
# Appropriately labels the data set with descriptive variable names.
names(totalSet) <- gsub("\\(\\)", "",names(totalSet))
names(totalSet) <- sub("activityLabel", "activityName", names(totalSet))
names(totalSet) <- sub("^t", "Time", names(totalSet))
names(totalSet) <- sub("^f", "Frequency", names(totalSet))
names(totalSet) <- sub("Acc", "Accelerometer", names(totalSet))
names(totalSet) <- sub("Gyro", "Gyroscope", names(totalSet))
```

5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject:

```{r}
# Creates a second, independent tidy data set with the average of each variable for each activity and each subject.
meltedDataSet <- melt(totalSet, id=c("subjectId","activityName"))
tidyDataSet <- dcast (meltedDataSet, subjectId + activityName ~ variable, mean)

# Removing data not longer needed
rm (meltedDataSet)

# Adding Average to variables
names(tidyDataSet)[3:68] <- paste0("Average",names(tidyDataSet)[3:68])

str(tidyDataSet)
write.table (tidyDataSet,"./data/tidyDataSet.txt", row.name=FALSE)
```


